---
title: "Exploring Some Activity Monitoring Data"
author: "Jesse Harris"
date: "March 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Assuming that the activity.csv file exists within the working directory, we can
load the data as follows:

```{r}
activity <- read.csv("activity.csv")
```

Let's see what the data looks like.

```{r}
str(activity)
```

Looks okay, but we should convert the dates from factor to date class.

```{r}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```

Okay, our data should be ready for analysis.

```{r}
str(activity)
```

Looks good. There are some NAs, but we'll ignore them for now.

## What is mean total number of steps taken per day?

First, let's calculate the total number of steps taken each day.

```{r}
df1 <- aggregate(steps ~ date, data = activity, sum)
head(df1)
```

Next, let's plot this as a histogram to get a feel for the distribution of the
data. We'll use the ggplot2 library to do this.

```{r}
library(ggplot2)
p <- ggplot(df1,
            aes(x = steps)) +
        geom_histogram(binwidth = 1000,
                       col = "blue",
                       fill = "blue",
                       alpha = 0.4,
                       breaks=seq(0, 22000, by=1000))
print(p)
```

Obviously, there are a lot of days where the number of steps is between 10,000
and 11,000. Let's see what the mean and median values work out to exactly.

```{r}
print(mean(df1$steps))
print(median(df1$steps))
```

## What is the average daily activity pattern?

First, let's figure out the average number of steps taken for each time
interval across all days.

```{r}
df2 <- aggregate(steps ~ interval, data = activity, mean)
```

To make sure this worked, let's have a look at a section of the data,
specifically, the average values for 7-8 AM.

```{r}
df2[85:97,]
tail(df2, length(10))
```

It seems to look okay. Let's make this data into a time series plot.

```{r}
p2 <- ggplot(df2, aes(interval, steps)) +
        geom_line()
print(p2)
```

So which 5-minute interval contains, on average, the maximum number of steps?

```{r}
df3 <- df2[order(df2$steps,decreasing=T),]
head(df3)
```

Looks like it's 08:35:00 to 08:39:59, which seems to match what we saw in our
time series plot above.

## Imputing missing values

How many of the rows in the dataset contain NA values?

```{r}
totalna <- sum(!(complete.cases(activity)))
print(totalna)
```

So 2304 rows. For interest's sake, let's check what percentage of the
values that represents.

```{r}
totalna/nrow(activity)*100

```

So let's impute the missing values to see if that will substantially change the
results of previous analyses. The method we'll use is to replace NA values with
the mean value, across all days, for that same time interval. The mean value
will be taken from the df2 data frame, created earlier, which contains all the
intervals and the mean steps value for each.

```{r}
activity2 <- activity
activity2$steps[is.na(activity2$steps)] <- df2$steps[match(activity2$interval,
                                                         df2$interval)]
```

The warning looks ugly, but I can assure you the code did what it needed to do.
:) Earlier when we did ran str against the activity data frame, the first steps
values were all NAs. Let's check it now.

```{r}
str(activity2)
```

So the replacement of NA values appears to have worked. Let's double check to
make sure there are no NAs left in the data set.

```{r}
totalna <- sum(!(complete.cases(activity2)))
print(totalna)
```

Okay, now let's generate our histogram again and see if there is any difference.
First we'll again calculate the total number of steps each day.

```{r}
df3 <- aggregate(steps ~ date, data = activity2, sum)
```

Now let's re-print our original histogram, and then we'll plot the new
histogram below for comparison.

```{r}
print(p)

p3 <- ggplot(df3,
            aes(x = steps)) +
        geom_histogram(binwidth = 1000,
                       col = "blue",
                       fill = "blue",
                       alpha = 0.4,
                       breaks=seq(0, 22000, by=1000))
print(p3)
```

So the overall shape is the same, however the number of days with 10000-11000
steps has increased significantly.

What about the mean and median values? Here are the mean and median with the NAs
left in.

```{r}
print(mean(df1$steps))
print(median(df1$steps))
```

And here are the new mean and median with the imputed data.

```{r}
print(mean(df3$steps))
print(median(df3$steps))
```

Wait, how come the mean and median ended up the same? Well, the original data
included six days with no observations. Every interval for those days was filled
with the same imputed data, that is, the mean for each interval, and therefore
all those days ended up having the same total number of steps.

This also explains why the number of days with step counts between 10,000
and 11,000 increased signifantly in our new histogram.

## Are there differences in activity patterns between weekdays and weekends?

Let's create a new factor variable that identifies whether a particular
observation occurred on a weekday or weekend.

```{r}
library(dplyr)
activity3 <- mutate(activity2, day1 = weekdays(activity2$date))
# or use base transform
activity3 <- mutate(activity3, day1 = gsub("S(atur|un)day", "weekend", day1))
filter(activity3, date == "2012-10-27")
```